---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Analysis of control experiments
This notebook looks at the impacts of movement variables and demography on population outcomes over 200 year runs.

## Load libraries and pre-process the data
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(janitor)

## Read and pre-process data
setwd("~/Documents/code/wasps/experiments")
wasps_control <- read.csv("wasps CONTROL-EXPERIMENT-070322-table.csv", skip = 6)
```

The initial experimental runs did not calculate some useful totals and the names used by NetLogo behaviour space are pretty strange, so some clean up needed.

```{r}
wasps_sel <- wasps_control %>%
  clean_names() %>%
  rename(run_number = x_run_number, t = x_step) %>%
  select(run_number, d_mean, p_ldd, birth_rate, t, 
         total_pop, prop_occupied, 
         sum_item_1_pops_of_the_habitable_land, 
         sum_item_2_pops_of_the_habitable_land) %>%
  rename(type1_pop = sum_item_1_pops_of_the_habitable_land,
         type2_pop = sum_item_2_pops_of_the_habitable_land) %>%
  mutate(gm_pop = type1_pop + type2_pop,
         wild_pop = total_pop - gm_pop)
```

## Determine the initial populations in each run
```{r}
wasps_start_pop <- wasps_sel %>%
  filter(t == 0) %>%
  select(run_number, total_pop) %>%
  rename(pop_0 = total_pop)
# and join to the data and calculate populations relative to start pop
wasps_sel <- wasps_sel %>%
  merge(wasps_start_pop, all.x = TRUE) %>%
  mutate(relative_pop = total_pop / pop_0)
```

## Determine final populations and add to data
For final pops we take populations over the last 25 years of each run

```{r}
final_pops <- wasps_sel %>% 
  filter(t >= 175) %>%
  group_by(run_number) %>%
  # and calculate final population and occupancy over that period
  summarise(pop_f = mean(relative_pop),
            occupancy_f = mean(prop_occupied))
# add to the data
wasps_sel <- wasps_sel %>%
  merge(final_pops)
```

### Write the tidied data out
```{r}
wasps_sel %>% write.csv("tidied_CONTROL-EXPERIMENT-070322-table.csv")
```

# Make some plots
## Population and occupancy time series
Facetted small multiples of population and occupancy time series fixing each variable in turn.

```{r}
plot_occ_pop_by_x <- function(fixed = "birth_rate", 
                              grid_vars = c("d_mean", "p_ldd"), 
                              val, fmt = "pdf", dpi = 600, 
                              width = 100, height = 75, units = "mm") {
  d <- wasps_sel %>%
    filter(.data[[fixed]] == val)
  xprint <- format(val, nsmall = 1)
  title <- paste("Population and occupancy by ", grid_vars[1], " and ", 
                 grid_vars[2], ", ", fixed, "=", xprint, sep = "") 
  fname <- paste("baseline_control_experiments/",
                 "pop_occ_by_", grid_vars[1], "_", grid_vars[2], "_", 
                 fixed, "_", xprint, ".", fmt, sep = "")
  
  g <- ggplot(d, aes(x = t, group = run_number)) + 
    geom_line(aes(y = relative_pop, colour = "Population"),
                lwd = 0.25, alpha = 0.25) +
    geom_line(aes(y = prop_occupied, colour = "Occupied area"), 
                lwd = 0.5, alpha = 0.25) +
    labs(x = "Time, generations", y = "Fraction of initial value", 
                colour = "Parameter") +
    facet_grid(reformulate(grid_vars[1], grid_vars[2]),
                labeller = label_both, as.table = FALSE) +
    ggtitle(title) +
    theme(text = element_text(size = 6, family = "sans"),
          panel.grid = element_line(size = 0.25, colour = "white"),
          axis.ticks = element_line(size = 0.25))

  print(g) # for the HTML output

  ggsave(fname, width = width, height = height, 
         units = units, dpi = dpi)
}
```

```{r}
for (x in unique(wasps_sel$birth_rate)) {
  plot_occ_pop_by_x(val = x)
}
for (x in unique(wasps_sel$d_mean)) {
  plot_occ_pop_by_x("d_mean", c("birth_rate", "p_ldd"), val = x)
}
for (x in unique(wasps_sel$p_ldd)) {
  plot_occ_pop_by_x("p_ldd", c("birth_rate", "d_mean"), val = x)
}
```


## Grid plots of final population and occupancy
These might be useful... unclear...

```{r}
grid_plot_data <- wasps_sel %>%
  group_by(birth_rate, d_mean, p_ldd) %>%
  summarise(mean_pop_f = mean(pop_f),
            mean_occ_f = mean(occupancy_f)) %>%
  ungroup()
```


```{r}
ggplot(grid_plot_data) + 
  geom_contour_filled(aes(x = birth_rate, y = d_mean, z = mean_pop_f)) +
  scale_fill_brewer(palette = "Reds") +
  facet_wrap(~ p_ldd) +
  ggtitle("Final mean population") +
  theme(text = element_text(size = 6, family = "sans"),
        panel.grid = element_line(size = 0.25, colour = "white"),
        axis.ticks = element_line(size = 0.25))
ggsave("baseline_control_experiments/final_mean_pop_by_birth_rate_d_mean_p_ldd.pdf",
       width = 100, height = 75, units = "mm", dpi = 600)

ggplot(grid_plot_data) + 
  geom_contour_filled(aes(x = birth_rate, y = d_mean, z = mean_occ_f)) +
  scale_fill_brewer(palette = "Reds") +
  facet_wrap(~ p_ldd) +
  ggtitle("Final mean occupancy") +
  theme(text = element_text(size = 6, family = "sans"),
        panel.grid = element_line(size = 0.25, colour = "white"),
        axis.ticks = element_line(size = 0.25))
ggsave("baseline_control_experiments/final_mean_occ_by_birth_rate_d_mean_p_ldd.pdf",
       width = 100, height = 75, units = "mm", dpi = 600)

```
