# Analysis of control experiments

## Load libraries

```{r}
library(dplyr)
library(ggplot2)
```

## Read and pre-process data

```{r}
setwd("~/Documents/code/wasps/experiments")
wasps.control <- read.csv("", skip=6)
```

Select the needed variables and determine total wild and GM populations and do some renaming.

```{r}
wasps.sel <- wasps.control %>%
  select(X.run.number., d.mean, p.ldd, birth.rate, X.step., 
         total.pop, prop.occupied, 
         sum..item.1.pops..of.the.habitable.land, 
         sum..item.2.pops..of.the.habitable.land) %>%
  rename(type1.pop = sum..item.1.pops..of.the.habitable.land,
         type2.pop = sum..item.2.pops..of.the.habitable.land) %>%
  mutate(gm.pop = type1.pop + type2.pop,
         wild.pop = total.pop - gm.pop)
```

Determine initial populations, and add to the data

```{r}
wasps.start.pop <- wasps.sel %>%
  filter(X.step. == 0) %>%
  select(X.run.number., total.pop) %>%
  rename(initial.pop = total.pop)
```

Use the initial population to rescale the data as a proportion of initial.

```{r}
wasps.sel <- wasps.sel %>%
  merge(wasps.start.pop, all.x = TRUE) %>%
  mutate(relative.pop = total.pop / initial.pop)
```

## Make some plots
### Population and occupancy by R_mean
Function to make facet small multiples of population and occupancy time series for a given intrinsice rate of reproduction, r

```{r}
plot_occupancy_population_by_dmean_pldd <- 
  function(data, r, save = TRUE, fname = "pops_by_dmean_pldd_birth_rate_", fmt = "pdf") {
  d.to_analyse <- data %>%
    filter(birth.rate == r)
  g <- ggplot(d.to_analyse, aes(x = X.step., group = X.run.number.)) + 
    geom_line(aes(y = relative.pop, colour = 'Population'), lwd = 0.25) +
    geom_line(aes(y = prop.occupied, colour = 'Occupied area'), lwd = 0.5) +
    labs(x = 'Time, generations', y='Fraction of initial value', colour = 'Parameter') + 
    facet_grid(p.ldd ~ d.mean, labeller=label_both, as.table = FALSE) +
    ggtitle(paste(
      'Population and occupancy by d_mean and p_LDD, birth_rate=', 
      as.character(r), sep = ''))
  if (save) {
    fig_name <- paste(fname, as.character(r), '.', fmt, sep = '')
    ggsave(fig_name, device = fmt, width = 4, height = 3, units = "in", scale = 2)
  }
  g
}

for (r in unique(wasps.sel$birth.rate)) {
  plot_occupancy_population_by_dmean_pldd(
    wasps.sel, r, 
    fname = "pops_by_dmean_pldd_birth_rate_", fmt = "png")
}
```

### Population and occupancy by a given p_ldd

```{r}
plot_occupancy_population_by_lambda_rmean <- function(data, pldd) {
  d.to_analyse <- data %>%
    filter(p.ldd == pldd)
  ggplot(d.to_analyse, aes(x = X.step., group = X.run.number.)) + 
    geom_line(aes(y = prop.occupied, colour = 'Occupied area'), lwd = 0.5, alpha = 0.25) +
    geom_line(aes(y = relative.pop, colour = 'Population'), lwd = 0.25, alpha = 0.25) +
    labs(x = 'Time, generations', y = 'Fraction of initial value', colour = 'Parameter') + 
    facet_grid(birth.rate ~ d.mean, labeller = label_both, as.table = FALSE) +
    ggtitle(paste('Population and occupancy by mean_d and birth_rate, p_LDD=', as.character(pldd), sep = ''))
  fig_name <- paste(
    'pops_by_birth_rate_d_mean_pldd_', as.character(pldd), '.pdf', sep = '')
  ggsave(fig_name, device = 'pdf')
}

for (pldd in unique(wasps.sel$p.ldd)) {
  plot_occ_pop_by_lambda_rmean(wasps.sel, pldd)
}


# 
# ## ----------------------------------------
# ## BOXPLOTS and VIOLIN PLOTS
# ## ----------------------------------------
# ## BY birth.rate
# d.to_analyse <- wasps.sel %>%
#   filter(X.step. >= 40, X.step. <=60) %>%
#   mutate(pldd = as.factor(p.ldd))
# 
# ggplot(d.to_analyse, aes(x = pldd)) + 
#   geom_boxplot(aes(y = prop.occupied, fill = 'Occupied area'), lwd = 0.25) +
#   geom_boxplot(aes(y = relative.pop, fill = 'Population'), lwd = 0.25) +
#   labs(x = 'Proportion long distance dispersal', y = 'Fraction of initial value', colour = 'Parameter') + 
#   facet_grid(birth.rate ~ d.mean, labeller = label_both, as.table = FALSE) +
#   ggtitle('Variation from generation 40-60')
# 
# ggplot(d.to_analyse, aes(x = pldd)) + 
#   geom_violin(aes(y = prop.occupied, fill = 'Occupied area'), alpha = 0.65, lwd = 0) +
#   geom_violin(aes(y = relative.pop, fill = 'Population'), alpha = 0.65, lwd = 0) +
#   labs(x = 'Proportion long distance dispersal', y = 'Fraction of initial value', fill = '') + 
#   facet_grid(birth.rate ~ d.mean, labeller=label_both, as.table=FALSE) +
#   ggtitle('Variation from generation 40-60')
# 
# ## BY P.LDD
# d.to_analyse <- wasps.sel %>%
#   filter(X.step. >= 40, X.step. <= 60) %>%
#   mutate(Rmean = as.factor(birth.rate))
# 
# ggplot(d.to_analyse, aes(x = Rmean)) + 
#   geom_boxplot(aes(y = prop.occupied, fill='Occupied area'), lwd=0.25) +
#   geom_boxplot(aes(y = relative.pop, fill='Population'), lwd=0.25) +
#   labs(x = 'Mean population growth rate', y='Fraction of initial value', colour = 'Parameter') + 
#   facet_grid(p.ldd ~ d.mean, labeller = label_both, as.table = FALSE) +
#   ggtitle('Variation from generation 40-60')
# 
# ggplot(d.to_analyse, aes(x = Rmean)) + 
#   geom_violin(aes(y = prop.occupied, fill = 'Occupied area'), alpha = 0.65, lwd = 0) +
#   geom_violin(aes(y = relative.pop, fill = 'Population'), alpha = 0.65, lwd = 0) +
#   labs(x = 'Mean population growth rate', y = 'Fraction of initial value', fill = '') + 
#   facet_grid(p.ldd ~ d.mean, labeller = label_both, as.table = FALSE) +
#   ggtitle('Variation from generation 40-60')
# 
